---
title: "Lab 2: Environmental Data"
author: "Tutorial by Fiona Spooner, following tutorials by Richard Pearson for courses at UCL"
output: html_document
---

The following video by Richard Pearson is relevant for this Lab:

https://www.youtube.com/watch?v=8inEr1c2UmE&list=PLKYTvTbXFuChaoF-L-1e9RzCagdLPQcCU&index=3

Additionally, a lot of the material in these practicals are adapted from Hijmans and Elith (2011) - https://cran.r-project.org/web/packages/dismo/vignettes/sdm.pdf

###Aims

* To view, clip, format and export environmental resource data


First we need to install the packages needed for this practical:
```{r, warning=F, message=F, eval=F}
install.packages("raster")
install.packages("maps")
install.packages("mapdata")
```

```{r, warning=F, message=F}
library(raster)
library(maps)
library(mapdata)
```

And to create a folder to put the environmental variables in - you will need to create the folder first before setting it as your working directory for this lab. This folder should be separate to the one you saved the species locations in.

```{r, eval=F}
setwd("C:/Users/Fiona/Desktop/SDM_Course/Env_Data") 

```

###Downloading environmental data

Now we can download the bioclim data directly from Worldclim. The environmental data are made up of 19 bioclimatic variables, the details of which you can read about here: http://www.worldclim.org/bioclim

The getData() can download the bioclim data at a variety of resolutions.  Valid resolutions are 0.5, 2.5, 5, and 10 (minutes of a degree). In the case of res=0.5, you must also provide a lon and lat argument for a tile; for the lower resolutions global data will be downloaded.

If you are a species of fairly restricted range you may be best of downloading the 0.5 resolution data but if the species is more broadly distributed it is probably best to use the 2.5 resolution.

```{r}

env = getData("worldclim", var="bio", res=2.5)

```

We can plot the environmental data, here are the first and twelfth layers, annual mean temperature and annual precipitation, respectively.

```{r}

bioclim_names<-c("Annual Mean Temp", "Mean Diurnal Range", "Isothermality", "Temp Seasonaliity", "Max Temp Warmest Month", "Min Temp Coldest Month", "Temp Annual Range", "Mean Temp Wettest Quarter","Mean Temp Driest Quarter", "Mean Temp Warmest Quarter", "Mean Temp Coldest Quarter", "Annual Precip", "Precip Wettest Month", "Precip Driest Month", "Precip Seasonality", "Precip Wettest Quarter", "Precip Driest Quarter", "Precip Warmest Quarter", "Precip Coldest Quarter")

names(env)<-bioclim_names

plot(env[[1]], main=names(env[[1]]))

plot(env[[12]], main="Annual Precipitation")

```


The data is currently at a global level, it needs to be cropped to the area of interest so that Maxent does not try to run the model for the whole world.

```{r, echo=F}
capg<-read.csv("capra_locs.csv")
```

```{r}

e<-extent(min(capg$lon)-1,max(capg$lon)+1,min(capg$lat)-1,max(capg$lat)+1)    
#should be in the order xmin,xmax,ymin,ymax
envcrop <- crop(env,e)

```


As you can see it is unlikely that annual mean temperature will vary from -200 - 300 degrees, the data are shown like this because the values are stored as integers (the real values multiplied by 10), so to see the 'real' values we can divide by 10.

```{r}
envcrop2<-envcrop/10

```


Now we can plot this cropped area, with 'real' values, with the location points on top.

```{r}

plot(envcrop2[[1]], main="Annual Mean Temperature")
map('worldHires',xlim=c(min(capg$lon)-1,max(capg$lon)+1), 
    ylim=c(min(capg$lat)-1,max(capg$lat)+1), fill=F, add=T)
points(capg$lon, capg$lat, pch="+")

```

Now you need to pick which of the bioclimatic variables you want to include in the Maxent model, try and pick ones which are most ecologically relevant to your species. Here I have picked the bioclimatic variables 1, 4, 11, 12, 14 and 19. They need to be exported as ascii files so that they can be used in Maxent.

```{r}
#listenv <- unstack(envcrop)

writeRaster(stack(envcrop2[[c(1,4,11,12,14,19)]]), names(envcrop2[[c(1,4,11,12,14,19)]]), 
            bylayer=TRUE, format='ascii', overwrite=T)

```

Now you have a folder with your environmental ascii layers in, which you can now use with your species locations to build a model in Maxent.




